Approximation as Differentiation: Notes

### 2025-05-27 notes/stability.tex

@@

Stability was invented by \citet{berry79} in an attempt to capture
sequentiality, and was also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien} (Chapter 12). Here, we are not
concerned with completeness properties considered in domain theory, so
we start with only partially ordered sets. Stable functions can be
defined as a certain subset of the set of monotone functions between
two partially ordered sets:

>>>>

Central to the concept of Galois slicing is the idea that
approximations of a value are a lattice, and that the functions
transporting approximations forwards and backwards form a Galois
connection between those lattices. We present a motivation for this
theory starting from the idea of qualitative information ordering from
Domain Theory. We argue that, with reference to differentiability,
monotone functions between certain partially ordered sets are
analogous to differentiable functions, and \emph{stable} functions
between those sets are analogous to functions with a reverse
derivative.

The stable function approach works, but mixes values and their
approximations into a single set. We use the families construction
over the category of lattices and Galois connections to separate
approximation from computation, and to provide a stepping stone to
executable models of Galois slicing.

\subsubsection{Domain Theory}

Domain theory is a method for giving denotational semantics to
programs that handle infinite data (i.e., functions,
streams). \emph{Domains} are special partially ordered sets where the
ordering denotes a relationship of qualitative information content: if
$x \sqsubseteq y$, then $y$ may contain more information than $x$. For
example, if $x$ and $y$ are functions, then $y$ may be defined at more
places in its domain than $x$.

Domains are typically closed under least upper bounds (\emph{lubs}) of
directed sets, meaning that any collection of elements that is
internally consistent has a ``completion'' that contains all the
information covered by the set. These directed lubs are used to give
semantics to recursive programs in terms of their finite
approximations.

Programs themselves are interpreted as monotone functions that
preserve directed lubs. The latter property is called
continuity. Monoticity captures the idea that if the input gets more
defined, then the output can get more defined. Continuity intuitively
expresses the idea that a function interpreting a program cannot
``look at all the approximations'' when determining the output, which
corresponds to the intuitive idea that a computable function cannot
look at a non-finite amount of input to produce an output.

For the purposes of Galois slicing, we are not interested in using
approximations to model computation on infinite objects, but rather to
use them for the related task of revealing how programs explore their
inputs for provenance tracking. Therefore, we start by only
considering certain partially ordered sets and monotone functions
between them.

\subsubsection{Forwards Slicing}

We now set up an analogy between certain partial orders with monotone
functions and manifolds and smooth functions.

In a partially ordered set $X$, for any $x \in X$ the set of elements
below $x$, $\downset{x} = \{ x' \mid x' \sqsubseteq x \}$, is itself a
partially ordered set. These approximations of $x$ we will think of as
``derivatives'' of $x$, and the whole set $\downset{x}$ as the
``tangent space''. Tangent spaces are vector spaces, since they are
linear approximations to curves at that point. In the partially
ordered setting, we take elements of $\downset{x}$ to be
approximations of processes defined at $x$. As we can add tangents, we
assume we can take meets of approximations in $\downset{x}$:

\begin{definition}
  A \emph{meet approximation space} is a partially ordered set $X$
  where every for every $x \in X$, $\downset{x}$ is a meet
  semilattice, with $x$ as the top element.
\end{definition}

Smooth functions have a ``pushforward'' derivative that takes tangents
at a point $x$ to tangents at the point $f(x)$. In the partially
ordered setting, we require that the restriction of a monotone
function $f$ to each $\downset{x}$ preserves meets:

\begin{definition}
  A \emph{morphism of meet approximation spaces} $f : X \to Y$ is a
  monotone function such that for all $x \in X$, the restriction
  $f_x : \downset{x} \to \downset{f(x)}$ preserves meets.
\end{definition}

\begin{lemma}
  Morphisms of meet approximation spaces are closed under identities
  and composition, forming a category. FIXME: with products?
  functions? sums?
\end{lemma}

\begin{example}
  FIXME: various forms of OR:
  \begin{itemize}
  \item strict and left-short-circuiting
  \item parallel or
  \item gustave
  \end{itemize}
\end{example}

\begin{example}
  An example that preserves meets, but isn't stable (from Amadio and
  Curien just before Lemma 12.2.3, originally due to Berry): Define
  $f : D \to \{\bot, \top\}$, where:
  \begin{displaymath}
    D = \bot \sqsubseteq \cdots \sqsubseteq n \sqsubseteq \cdots \sqsubseteq 1 \sqsubseteq 0
  \end{displaymath}
  as $f(\bot) = \bot$ and $f(n) = \top$. This is monotone, and
  preserves meets in every $\downset{x}$. However, there is no
  ``best'' (i.e., least) input that gives us any finite output. To
  rectify this, we will further assume that our functions are stable.
\end{example}

\subsubsection{Backwards Slicing}

@@

Stability was invented by Berry \cite{berry86} in an attempt to capture
sequentiality, and were also instrumental in the discovery of Linear
Logic via Coherence Spaces \cite{girard}. A textbook description of
stable functions in the context of domain theory is given by Amadio
and Curien \cite{amadio-curien} (Chapter 12). Here, we are not
concerned with completeness properties considered in domain theory, so
we start with only partially ordered sets. Stable functions can be
defined as a certain subset of the set of monotone functions between
two partially ordered sets:
